---
title: "Application of poisson mash and mash to single cell cytokines data (Ly6C- monocytes)"
author: "Yusha Liu"
header-includes:
date: "2020-9-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Summary
We apply both poisson mash and mash to a subset of Ly6C- monocytes from the cytokines data, treated by one of 25 selected conditions from the second batch. We apply both methods to the $6,794$ genes with at least $100$ reads across all cells, so the count data matrix $X$ is $6,794 \times 25$, where $X_{jr}$ represents the total read counts of gene $j$ in all cells from condition $r$. 

Recall the [barplot and enrichment-depletion plot](https://github.com/stephenslab/single-cell-cytokines/blob/master/output/results_by_cell_type/Ly6C-_Monocytes_barplots.pdf) which display single cell-specific loadings grouped by conditions, based on topic modeling of single cell data from Ly6C- monocytes (the number of topics $K=12$).
```{r warning=FALSE, message=FALSE}
library(pheatmap)
library(gridExtra)

cols.trt <- readRDS("output/Ly6C-/condition_colors.Rds")
trts <- names(cols.trt)

```



### Model
We assume $X_{jr} \sim Pois(s_r \lambda_{jr})$, and $\log(\lambda_{jr}) = \mu_j + \beta_{jr} + \eta_{jr}$, where $s_r$ accounts for the different sequencing depth of each condition, $\mu_j$ is the gene-specific mean on the log scale shared by all conditions, $\beta_{jr}$ models the condition-specific treatment effect, and $\eta_{jr}$ models the random effect. We are mainly interested in estimation and inference on $\beta_{jr}$. The following priors are assumed in our model:
\begin{align}
& \beta_j  =  \sum_{h, l} \pi_{h, l} \: MVN(0, \: w_l U_h) + \sum_{g, l} \pi_{g, l} \: MVN(0, \: w_l u_g u'_g) \\
\text{where} \quad & \sum_{h, l} \pi_{h, l} + \sum_{g, l} \pi_{g, l} = 1 \\
& \eta_j \sim MVN(0, \: \tau_j^2 I_R). \\
\end{align}



### ED step
To estimate the data-driven prior covariances $U_h$ and $u_g u'_g$, we fit the poisson mash model without scaling parameters $w_l$ to a subset of genes that are differentially expressed across conditions, which are identified based on a conditional multinomial goodness-of-fit test. 
```{r warning=FALSE, message=FALSE}
## load in the ED results
fit.ed <- readRDS("output/Ly6C-/ed_orig_data.Rds")

## running time per iteration in ED
fit.ed$runtime/length(fit.ed$ELBO)

## estimated weights for different data-driven prior covariances
names(fit.ed$pi) <- c(names(fit.ed$Ulist), names(fit.ed$ulist))
round(fit.ed$pi, 3)

## look at the ELBO
plot(fit.ed$ELBO, xlab = "niter", ylab = "elbo", main="fitting ED step")

```



### Estimate of data-driven covariances from ED step
```{r warning=F}
par(mfrow=c(1,1))

### Estimate of eigenvectors of the data-driven covariance initialized by rank-5 PCA
eig.tPCA <- eigen(fit.ed$Ulist[[1]])
pve.tPCA <- eig.tPCA$values/sum(eig.tPCA$values)

for (k in 1:4){
  v <- eig.tPCA$vectors[,k]
  
  barplot(v/v[which.max(abs(v))], names = names(cols.trt), cex.names = 0.4,
          las = 2, main = paste0(names(fit.ed$Ulist)[1], ": weight ", round(fit.ed$pi[1],3), 
                                 ":\nEigenvector ", k, " (pve = ", round(pve.tPCA[k],3), ")"), col = cols.trt)
}


### Estimate of eigenvectors of rank-1 data-driven covariance matrices
for (k in 1:5){
  v <- fit.ed$ulist[[k]]
  
  barplot(v/v[which.max(abs(v))], names = names(cols.trt), cex.names = 0.4,
          las = 2, main = paste0(names(fit.ed$ulist)[k], ": weight ", round(fit.ed$pi[2+k],3)), col = cols.trt)
}

```



### Running poisson mash
We then look at the poisson mash fit to the entire dataset. 
```{r warning=FALSE, message=FALSE}
## load in the poisson mash fit results
res <- readRDS("output/Ly6C-/pois_mash_fit_orig_data.Rds")
res$runtime

## look at ELBO
plot(res$ELBO, xlab = "niter", ylab = "elbo", main="Fitting poisson mash with random effects")

## look at distribution of random effect size (in terms of standard deviation on the log scale)
quantile(sqrt(res$tau2), seq(0.05, 0.95, 0.05))

## look at the estimated weights of prior covariance matrices
wlist <- c(1e-3, 2.5e-3, 5e-3, 1e-2, 2e-2, 4e-2, 8e-2, 0.16, 0.32, 0.64, 1, 2, 4, 8, 16, 25, 36, 50)

pi.mat <- matrix(res$pi, ncol=length(wlist), byrow=TRUE)
rownames(pi.mat) <- c(names(fit.ed$Ulist), paste0("e_", 1:25), names(fit.ed$ulist))
colnames(pi.mat) <- paste0("w=", wlist)

round(rowSums(pi.mat),3)

pheatmap(pi.mat, cluster_rows=FALSE, cluster_cols=FALSE, fontsize_row=5, fontsize_col=6, 
         main="Estimated weights of prior covariances in poisson mash fit")

```



### Comparison to fixed effects-based mash
We calculate posterior summaries of $\delta_{jr} = \beta_{jr} - \sum_{r=1}^R \beta_{jr}/R$ from the poisson mash fit, i.e., the condition-specific deviation from the mean across all conditions, and compare to the results from fixed effects-based mash. In particular, we fit a linear fixed effects model separately to each gene $j$, using the single-cell level normalized gene expression data as the response and the condition assignment of each cell as the predictor, to obtain $\hat{b}_{jr}$ and $\hat{s}_{jr}$ for each gene $j$ and condition $r$. When fitting this model, we assume that the Gaussian observation noise is constant over conditions for each gene $j$. We then feed $\hat{b}_{jr}$ and $\hat{s}_{jr}$ to mash without common baseline. 
```{r warning=FALSE, message=FALSE}
## load in the posterior summaries from poisson mash fit
post <- readRDS("output/Ly6C-/pois_mash_posterior_orig_data.Rds")

## load in the mash fit
mash.fit <- readRDS("output/Ly6C-/mash_fit_orig_data.Rds")

## the number of significant genes identified by poisson mash 
idx.pois.mash <- which(apply(post$lfsr, 1, min) < 0.05)
length(idx.pois.mash)

## the number of significant genes identified by mash
idx.mash <- mashr::get_significant_results(mash.fit)
length(idx.mash)

## the number of genes that are identified by both 
sum(names(idx.mash) %in% names(idx.pois.mash))

## the number of genes that are identified only by poisson mash 
pois_mash_only <- idx.pois.mash[!names(idx.pois.mash) %in% names(idx.mash)]
length(pois_mash_only)

## the number of genes that are identified only by mash 
mash_only <- idx.mash[!names(idx.mash) %in% names(idx.pois.mash)]
length(mash_only)

```



